Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-09-17 paint-contour-tree.tex
modified authors
updating upper bounds

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\author{
  Seshadhri Comandur
  \and
  Kashyap Dixit
  \and
  Benjamin Raichel
}

\title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
%\date{}

@@

\pagebreak
\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Lower Bound}

In this section we prove a lower bound on the time it takes to compute a merge tree.

@@

First we setup some basic terminology.
Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.

\begin{definition}
 A coloring of a tree $T$ is any function $\chi:L(T)\times E \rightarrow [n]$, where $L(T)$ is the set of leaves of $T$ and $n = |L(T)|$.  
 We say a coloring is valid if
 \begin{itemize}
  \item The restriction $\chi:L(T) \rightarrow [n]$ is a bijection.
  \item Let $e$ be any edge in $E$ and let $u$ be its higher endpoint, then $\chi(e) \in \chi(L(T_u))$. 
 \end{itemize}
 We consider each internal vertex as having two colors, namely the colors of the two edges for which it is a lower endpoint.
\end{definition}

For a vertex $v\in T$, let $H_v$ be the subset of 
vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $\chi(L(T_v))$.
We will refer to $H_v$ as the \emph{heap} at $v$.
Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.

\Ben{The following lemma should be immediate once previous section is written properly.}
\begin{lemma}
\label{lem:algCost}
NEEDS TO BE WRITTEN:
Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
of this section proving that this running time is optimal.

To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

\begin{definition}
\label{def:path}
 Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
 is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
 For $p\in P(T)$, we use $|p|$ to denote its length and $r_p$ to denote the lowest vertex in $p$, which we call the \emph{root} of $p$.

 We define the cost of a particular path decomposition, $P(T)$, as $f_{path}(P(T)) = \sum_{p\in P(T)} |p| \log |p|$.
\end{definition}

Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub.

@@

\subsection{Lower Bound by Path Decomposition}

\subsection{Upper Bound by Path Decomposition}
\Ben{This section is a work in progress.}

From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
which we will call the maximum path decomposition (defined below).
Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
%using the observations about path decompositions from 
%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.

Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.

\subsubsection{Heavy and light paths}

Before describing the maximum path decomposition, we introduce the notion of heavy and light paths 
(which will help motivate what the decomposition should be).

Recall we are ultimately trying to show $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ for some path decomposition $P=P(T)$, 
or more specifically $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$.  In particular, if we could 
show for each path $p\in P$ that $\sum_{v\in p} \log |H_v|= |p|\log|p|$ then we would be done.
In other words we can think of each path as having a budget of $|p|\log|p|$, and ideally the budget 
of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
we will divide paths into two categories, called heavy and light, which as we show below in \Lem{pathBounds} roughly corresponds 
to whether a path can pay for itself or not.

\begin{definition}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
 For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and we say it is \emph{heavy} otherwise.
\end{definition}

\begin{observation}
\label{obs:decrease}
 Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
 $w$ to the root has one less vertex than the path from $v$ to the root.
 In particular, we have the following more general property.
 Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
 Then $|H_v| \leq |H_u| + d(u,v)$. 
\end{observation}

The following can be thought of as a generalization of the above observation, and will be useful in later sections.

\begin{lemma}
\label{lem:adjacent}
 Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$ in $\pathTree(T)$.
 For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
 Also, let $H_0$ denote the heap at the leaf of $p$.

\begin{lemma}
\label{lem:pathBounds}
 Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
 Then, 
 \begin{compactenum}
  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
 \end{compactenum}
\end{lemma}
\begin{proof}
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
 As such, 
 \[
 \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|).
 \]
 If $p$ is a heavy path then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
\end{proof}

For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
Our eventual goal will be to show that by choosing the right path decomposition, this excess 
can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.

\subsubsection{The decomposition and its main property}

\begin{definition}
 Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
 $T$ as follows.  For each internal vertex $v\in T$ add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
 where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
 The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
 we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
\end{definition}

We now prove that in a maximum path decomposition, when going between adjacent light paths, the heap size decreases by a constant fraction.

\begin{lemma}
\label{lem:geometric}
 Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the shrub $\pathTreeA$.  
 Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
\end{lemma}
\begin{proof}
 Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
 i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
 the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
 Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
 Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
the following definition.

@@

As the weight $w(p$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.

@@

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\author{
  Benjamin Raichel
  \and
  C. Seshadhri
}

\title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
\date{}

@@

\pagebreak
\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Leaf assignments and path decompositions}

In this section, we set up a framework to analyze the time taken to compute a merge tree.

@@

Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.
We use $A(v)$ to denote the set of ancestors of $v$.
For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.

\begin{definition}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
 one from the left and one from the right subtree of $v$.
\end{definition}

For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
The following lemma should justify these technical definitions.

\begin{lemma}
The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
% where $T$ is the merge tree of the given input, 
%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

\begin{proof} \Sesh{add}

\end{proof}

%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
%We will refer to $H_v$ as the \emph{heap} at $v$.
%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
%
%\begin{lemma}
%\label{lem:algCost}
%NEEDS TO BE WRITTEN:
%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
%\end{lemma}
%
%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
%of this section proving that this running time is optimal.

We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
containing a leaf.
% Then in the following two subsections 
%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

\begin{definition}
\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.

 The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
\end{definition}

Fix $\chi(T)$, which fixes all the heaps $H_v$. We choose a special path decomposition, which can be related
to the running time of the algorithm. This is defined best as a subset of edges in $T$ such that each internal
vertex has degree exactly $1$. This naturally gives as path decomposition.
For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
 where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.

Our main theorem is the following.

\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\cost(\pmax(T)))$.
\end{theorem}

\subsection{Shrubs, heavy paths, and light paths}

The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
The shrub gives a handle on the heaps of a path.
We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
We use $H_p$ to denote the $H_{b(p)}$.
% and denote it by $\pathTree(T)$.

\begin{lemma}
\label{lem:adjacent}
 Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
\end{lemma}

\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
Consider $v \in \bigcup_{i} H_i$ that lies below
$b(p)$ in $T$. Note that $v \in H_p$ and has only one color in $\bigcup_{i} L(T_(b(q_i))) \cup \ell(p)$.
Since the latter sets are all disjoint, any $u \in H_i \cap H_j$ must lie in $p$.
Any $u \in p$ can be in at most two of the $H_i$'s. Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
\ignore{
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
 Putting these together gives
 \[
  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
 \]
}
\end{proof}

We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
This is unfortunately not true, which is why we divide paths into two categories.

\begin{definition}
 For $p\in P(T)$, $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{heavy} otherwise.
\end{definition}

\ignore{
\begin{observation}
\label{obs:decrease}
 Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
 $w$ to the root has one less vertex than the path from $v$ to the root.
 In particular, we have the following more general property.
 Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
 Then $|H_v| \leq |H_u| + d(u,v)$. 
\end{observation}

The following can be thought of as a generalization of the above observation, and will be useful in later sections.
}

\begin{lemma}
\label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_p|)$.
% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
% Then, 
% \begin{compactenum}
%  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
%  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
% \end{compactenum}
\end{lemma}
\begin{proof} \ignore{
 By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
}
For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
 If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
 $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
\end{proof}

\ignore{
For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
Our eventual goal will be to show that by choosing the right path decomposition, this excess 
can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
}

We have a claim that also bounds heap sizes for special set of paths.

\begin{claim} \label{clm:above} Let $Q$ be a set of incomparable nodes in the shrub $P(T)$ (which are paths in $T$).
Suppose all paths in $A(Q)$ are heavy.
Then $\sum_{q \in Q} |H_q| \log |H_q| = O(\sum_{p \in A(Q)} |p|\log |p|)$.
\end{claim}

\begin{proof} Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.

Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
Each such $v$ must occur in some path $p \in A(Q)$. Thus,
$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
We apply the bound of \Lem{pathBounds} to complete the proof.
\end{proof}

Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
is light. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.

\begin{lemma} \label{lem:light} $\sum_{p \in L} \sum_{v \in p} \log |H_v| = O(\sum_{p \in L'} |H_p| \log |H_p|)$.
\end{lemma}

With this lemma, the proof of \Thm{runtime} is straightforward.

\begin{proof} (of \Thm{runtime}) We split into heavy and light paths.
We have $\sum_{v \in T} \log |H_v| = \sum_{p \in L} \sum_{v \in p} \log |H_v| + \sum_{p \notin L} \sum_{v \in p} \log |H_v|$.
For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in L'} |H_p| \log |H_p|)$.
Now, we apply \Clm{above} to the set $L'$, noting that it satifies the claim conditions and that $A(Q) \cap L = \emptyset$.
Hence, the first term is $O(\sum_{p \notin L} |p|\log |p|)$.
The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
$\sum_{p \notin L} |p|\log |p| \leq \cost(\pmax(T))$.
\end{proof}

The main challenge is proving \Lem{light}, which we defer to the next section.

\ignore{
\begin{definition}
 Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
 $T$ as follows.  
 The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
 we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
\end{definition}
}

\subsection{Upper Bound by Path Decomposition}

\ignore{
From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
which we will call the maximum path decomposition (defined below).
Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
%using the observations about path decompositions from 
%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.

Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
}

We focus on $\pmax(T)$, and prove the key property of this path decomposition.

\begin{lemma}
\label{lem:geometric}
 Let $p$ be the parent of $q$ in the shrub $\pmax(T)$.  
 Then $|H_q|\leq \frac{3}{5} |H_p|$. 
\end{lemma}
\begin{proof}
 Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
 i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
 the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
 Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
 Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

\ignore{
Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
the following definition.

We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
}

@@

@@

As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.

@@

\section{Lower Bound by Path Decomposition}

\appendix

\section{Old stuff}
